Probing heavy ion collisions using quark and gluon jet substructure

### 2018-02-21 qg_ML.tex
YT updated draft.

For a given jet flavor, i.e. quark or gluon jets, one can also study with various aforementioned methods how the medium modifies jets by comparing vacuum jets with quenched jets in the context of proton-heavy ion jet discrimination. Figure~\ref{fig:ROC_qq_gg} shows the ROC curves for discriminating quark jets (left panel) and gluon jets (right panel) in pp and PbPb collisions using physics-motivated variables, jet images and telescoping deconstruction. We see huge difference between jets in pp and \jwpy-simulated PbPb collisions. All the methods perform significantly better in the task of identifying if a jet is quenched or not, compared to quark gluon discrimination. Therefore we plot the ROC curves using the log scale in the $y$-axis. We observe that the increase of the pixel multiplicity is a key feature in identifying quenched jets, with the DCNN performs slightly better than the MLP. The better performance in the proton-heavy ion jet discrimination for gluon jets (right panel of Figure~\ref{fig:ROC_qq_gg}) suggests that gluon jets are modified more significantly than quark jets due to medium-induced quenching.

From our studies of the ROC curves for comparing quark jets and gluon jets, as well as comparing vacuum jets and quenched jets, we begin to qualitatively understand the effects of quenching as implemented in the \jwpy Monte Carlo. Quenched jets end up with significantly more jet constituents which are distributed away from the jet axis as seen in the saturation of the TD classification performance from T1 to T3 order. This modification appears to affect gluon jets significantly more than quark jets in \jw since a gluon jet, with its larger particle multiplicity and wider parton shower, can all interact with the scattering centers leading to broader and more quenched jets. In the next section we will employ the use of Lund Diagrams to study the differences in quenched quark and gluon jets.

@@

Any jet observable on the market is designed as some function of the momentum and the distance scale that is used in the jet reconstruction. These two scales can be represented as $\ln (R/\Delta R)$ and $\ln z(\Delta R/R)$, where $z$ is the transverse momentum fraction of the softer subjet relative to sum of the two subjets, $\Delta R$ is the angular scale between the two subjets in a given branch and $R$ is the jet radius. Such a representation of a jet is known as the Lund Diagram~\cite{Salam:2016yht} and the prescription to create such a diagram for a given jet is as follows:
\begin{itemize}
	\item Cluster your jet with a jet R say $0.4$ and a particular jet algorithm, in our case anti-k$_{t}$
	\item Take the jet constituents and re-cluster with the C/A algorithm which combines the nearest particles with each other
	\item Walk backwards along the clustering tree and for each branch calculate the $z=p^{2}_{T}/(p^{1}_{T}+p^{2}_{T})$ and $\Delta R = \sqrt{\Delta \eta^2_{1, 2}+\phi^2_{1, 2}}$
	\item Fill the Lund diagram at each split
	\item Continue in walking backwards along the hard branch till you reach only one particle
\end{itemize}

@@

The beauty of a Lund diagram is its convenient separation of wide soft angle radiation close to the y-axis, collinear hard radiation along the diagonal and with contribution from hadronization and underlying event in a line parallel to the x-axis. These Lund diagrams have also been recently used in distinguishing between QCD and boosted heavy objects due to their unique feature of featuring jets of a constant mass along the main diagonal (opposite to the collinear hard radiation). We show the Lund diagrams for quark (left) and gluon (right)jets in pp (top) and PbPb (bottom) in Fig:~\ref{fig:Lund_full} where the z is calculated relative to each split. We see an enhanced region in PbPb collisions close to the y-axis in comparison to pp which corresponds to increased wide soft angle in quenched jets. Comparing left to right (as we discussed in Fig:~\ref{fig:quenching_discrimination}) one also sees gluon jets include a larger fraction of the soft radiation and these are susceptible to hadronization and underlying event effects. The hard collinear radiation region of the Lund diagram, that is dominated in pp is shown to reduce in its significance for quenched jets. The core of the jet remains similar between pp and PbPb, but the secondary sub-jet ultimately gets pushed to wider angles and to softer momenta. Such a picture is exactly what we discussed in Fig:~\ref{fig:comp-zg}-\ref{fig:comp_subjet_m} but this time, visible via the Lund diagram.

@@

One can now study the effect of grooming with the help of the Lund diagrams by rejecting the soft radiation ($z < 0.1$) that is typically groomed a way the softdrop method. Fig:~\ref{fig:Lund_bkg} shows the Lund diagrams for our jets in pp and PbPb  before the one which passes the soft-drop condition. We see that the soft, wide-angle radiation is isolated and the hard collinear core is essentially removed. The bottom right panel of Fig:~\ref{fig:Lund_bkg} with quenched gluon jets shows the maximum density of soft wide angle radiation. These soft jet constituents in the jet's periphery are precisely the ones that show up as a modification of the jet shape. Within the framework of this diagrammatic representation, one can now build up these observables and follow the harder branch leading from the first split. Fig:~\ref{fig:Lund_hard} shows the Lund diagrams after the soft branch (background) has been removed via the soft-drop procedure. We see that the vacuum structure starts to emerge in the soft-drop PbPb jets. However, soft-drop can not remove all the soft radiation within the hard branch in PbPb collisions as shown by the enhanced wide angle radiation. We understand this behavior as the correlated medium response that belongs to the jet but is often removed due to a statistical treatment of background in experiments.

@@

We have shown a systematic study of modification to the jet structure due to the effects of quenching in heavy ion collisions as implemented within the \jw MC generator. The medium modification is studied via the novel method of comparing and contrasting the classification of quark jets from gluon jets. We compare a MLP with physics motivated jet variables and a state of the art DCNN applied on jet images to a our method of Telescoping Deconstruction and find the classification performance worsens for quenched jets when compared with vacuum pp jets. All of the models studied perform consistently with each other giving us the confidence that the TD to third order is extracting fundamental fragmentation patters of a given jet.

DISCUSSION - THIS IS A VERY IMPORTANT PART

\section*{Acknowledgments}
The authors are grateful to Patrick Komiske and Eric Metodiev for collaborating on the development of telescoping deconstruction and contributions to the Appendix. The authors also thank Benjamin Nachman, Joern Putschke, Konrad Tywoniuk and Sevil Salur for helpful discussions. We thank Gavin Salam and the organizers of CERN jet workshop (2017) for bringing to our attention the concept of Lund Diagrams and their usage in studying jet modifications. YTC was supported by the LHC Theory Initiative Postdoctoral Fellowship under the National Science Foundation grant PHY-1419008. RKE acknowledges partial support from the NSF under Grant No.1067907 \& 1352081 and DOE (need to ask Joern).

>>>>

For a given jet flavor, i.e. quark or gluon jets, one can also study with various aforementioned methods how the medium modifies jets by comparing vacuum jets with quenched jets in the context of proton-heavy ion jet discrimination. Figure~\ref{fig:ROC_qq_gg} shows the ROC curves plotting the PbPb jet efficency as a function of the pp jet efficiency, for discriminating quark jets (left panel) or gluon jets (right panel) in pp and PbPb collisions using physics-motivated variables, jet images and telescoping deconstruction. We see huge differences between jets in pp and \jwpy-simulated PbPb collisions. All the methods perform significantly better in the task of identifying if a jet is quenched or not, compared to quark gluon discrimination. Therefore we plot the ROC curves using the log scale in the $y$-axis. We observe that the increase of the pixel multiplicity is a key feature in identifying quenched jets, with the DCNN performs slightly better than the MLP. The better performance in the proton-heavy ion jet discrimination for gluon jets (right panel of Figure~\ref{fig:ROC_qq_gg}) suggests that gluon jets are modified more significantly than quark jets due to medium-induced quenching.

From our studies of the ROC curves for comparing quark jets and gluon jets, as well as comparing vacuum jets and quenched jets, we understand qualitatively in more details the effects of quenching as implemented in the \jwpy Monte Carlo. Quenched jets end up with significantly more jet constituents which are distributed away from the jet axis, resulting in the saturation of the TD classification performance at the T1 order. This modification appears to affect gluon jets significantly more than quark jets in \jw since a gluon jet, with its larger particle multiplicity and wider parton shower, can all interact with the scattering centers leading to broader and more quenched jets. In the next section we will employ the use of another jet representation, the Lund Diagram, to illuminate the differences between quark and gluon jets, as well as jets in pp and PbPb collisions

@@

Each jet observable is defined as a function of the momenta and angles of jet constituents, which probes the entire jet formation with a single number. One can choose a set of jet observables to be a representation of jets, and we discussed previously several approaches all faithfully encode the jet information. In this section we will use the Lund diagram~\cite{Andersson1989} associated with the Cambridge/Aachen clustering tree structure as another jet representation. It has also been recently used in distinguishing between QCD and boosted heavy particle jets \cite{Salam:2016yht}. As we will see, the Lund diagram directly records the branching kinematics along the hard branches in the C/A clustering tree in terms of longitudinal momentum fraction $z$ of the soft branch and the angular separation $\Delta R$ between the two branches. The Lund diagram is then a two-dimensional histogram of variables $\ln (R/\Delta R)$ and $\ln z(\Delta R/R)$ where $R$ is the jet radius used in jet reconstruction. The variables are chosen such that the Lund diagram is uniform in the fixed strong coupling limit for jets in pp collisions because of the infrared structure of QCD.

Given a jet reconstructed using the anti-$k_{T}$ algorithm with $R = 0.4$, the procedure of creating the Lund diagram is as follows:
\begin{itemize}
	\item Recluster the jet using the C/A algorithm
	\item Start from the root of the tree and move along the hard branches. At each branching, construct the momentum fraction of the soft branch $z=\min(p_{T_1},p_{T_2})/(p_{T_1}+p_{T_2})$ and the angular distance between the two branches $\Delta R = \sqrt{\Delta \eta^2_{1, 2}+\phi^2_{1, 2}}$
	\item Fill in the Lund diagram for each branching
	\item Continue along the hard branches until reaching only one particle
\end{itemize}

@@

In the Lund diagram, wide angle, soft radiation populates close to the $y$-axis, while collinear, hard radiation is along the diagonal. Figure~\ref{fig:Lund_full} shows the Lund diagrams for quark (left) and gluon (right) jets in pp (top) and PbPb (bottom) collisions. We see a significantly enhanced region close to the $y$-axis in PbPb collisions which corresponds to increased wide angle, soft radiation in quenched jets. Also, gluon jets have a larger fraction of the soft radiation. On the other hand, the hard, collinear radiation region of the Lund diagram is shown to reduce in its significance for quenched jets. As we discussed in Figure~\ref{fig:comp_z}, \ref{fig:comp_rg} and \ref{fig:comp_subjet_m}, jet cores are also modified in PbPb collisions, and subleading subjets tend to have wider angles and softer momenta. Such picture emerges and is directly visible in the Lund diagram.

@@

We can also directly study the effect of soft-drop using the Lund diagram. Figure~\ref{fig:Lund_bkg} shows the Lund diagrams for quark and gluon jets in pp and PbPb collisions, with only the branchings before the one defining the $z_g$ and $r_g$ observables. That is, Figure~\ref{fig:Lund_bkg} corresponds to radiation that is dropped in the soft-drop procedure. Note the strict cutoff of $z < z_{\rm cut} = 0.1$ in the Lund diagrams. We see that a large portion of soft, wide-angle radiation is isolated from the hard, collinear core. The bottom right panel of Figure~\ref{fig:Lund_bkg} with quenched gluon jets shows the maximum density of soft, wide angle radiation. These soft jet constituents near the jet boundary are precisely the ones contribute to the modification of the jet shape in \jw \cite{KunnawalkamElayavalli:2017hxo}. Within the framework of this diagrammatic representation, one can also examine the harder branch after the soft-drop procedure and Figure~\ref{fig:Lund_hard} shows the corresponding Lund diagram. We see that the vacuum structure starts to emerge in the hard branch of soft-drop PbPb jets. However, soft-drop still can not remove all the soft radiation within the hard branch in PbPb collisions as shown by the enhanced wide angle radiation. This can be due to correlated medium responses but is often removed with a statistical treatment of background in experiments.

@@

We present a systematic study of modifications to quark and gluon jet structure in heavy ion collisions, taking the \jwpy Monte Carlo simulations as an example. Jet modifications are studied via the novel method of comparing and contrasting the classification of quark jets from gluon jets in different collision systems. We compare the performances of a MLP with physics-motivated jet variables, a state-of-the-art DCNN trained on jet images, and a newly developed method of telescoping deconstruction. We find that the quark gluon classification performance worsens for quenched jets. All the methods studied perform consistently, suggesting that the TD can extract fundamental jet fragmentation patterns. Through multiple methods and observables we observe consistently that the dominant feature in the \jwpy-simulated PbPb collisions is the increase of soft particle multiplicity throughout the jet region. This results in the lost of information in subleading subjets which is a characteristic feature of the \jw simulation.

The future of heavy ion jet physics program necessarily moves toward precise, simultaneous understanding of multiple jet substructure observables and their correlations. Therefore our work serves as an example how multivariate techniques can help illuminate the underlying mechanism of jet modifications. The telescoping deconstruction method suggests a framework of jet substructure observables for jet physics studies, which we show to give qualitatively new insights order-by-order in its observable organization. In the future the comparisons between theoretical calculations and simulations \cite{}, and with experimental measurements, will become standard practice to identify robust features of jet-medium interactions. Our work suggests a systematically improvable framework which will serve this purpose.

\section*{Acknowledgments}
The authors are grateful to Patrick Komiske and Eric Metodiev for collaborating on the development of telescoping deconstruction and contributions to the Appendix. The authors also thank Fr\'ed\'eric Dreyer, Benjamin Nachman, Joern Putschke, Konrad Tywoniuk and Sevil Salur for helpful discussions. We thank Gavin Salam and the participants of the CERN TH institute ``{\sl Novel tools and observables for jet physics in heavy ion collisions}" and the 5-th Heavy Ion Jet Workshop for bringing to our attention the Lund Diagram and its usage in studying jet modifications. YTC was supported by the LHC Theory Initiative Postdoctoral Fellowship under the National Science Foundation grant PHY-1419008. RKE acknowledges partial support from the NSF under Grant No.1067907 \& 1352081 and DOE (need to ask Joern).
